% +--------------------------------------------------+
% | Typeset this file to get the documentation.      |
% +--------------------------------------------------+
%
%% Copyright (C) 1998, 1999, 2000, 2006 Javier Bezos
%% All Rights Reserved
%% http://www.tex-tipografia.com
%%
%% This work may be distributed and/or modified under the conditions
%% of the LaTeX Project Public License, either version 1.3 of this
%% license or (at your option) any later version.  The latest version
%% of this license is in
%%     http://www.latex-project.org/lppl.txt
%% and version 1.3 or later is part of all distributions of LaTeX
%% version 2003/12/01 or later.
%%
%% This work has the LPPL maintenance status "maintained".
%%
%% This Current Maintainer of this work is Javier Bezos.
%%
%% This work consists of the files accents.tex and accents.sty.

\def\fileversion{1.3}
\def\docdate{May 12, 2006}

\documentclass{ltxguide}
\usepackage{accents}

\title{The \textsf{accents} Package\footnote{This 
package is currently at version \fileversion.}}

\author{Javier Bezos\footnote{For bug reports, comments and
suggestions go to
\texttt{http://www.tex-tipografia.com}.
English is not my strong point, so contact me when you find mistakes
in the manual.  Other packages by the same author: \textsf{tensind}
(tensorial indexes), \textsf{spanish} (for babel), \textsf{titlesec}
(to redefine sectioning macros), \textsf{dotlessi} (dotless i in any
math group).}}

\date{\docdate}

\newenvironment{sample}{\begin{quote}\small\begin{tabbing}%
  \hskip14pc\=\hskip6pc\\}
  {\end{tabbing}\end{quote}}

\DeclareMathAccent{\wtilde}{\mathord}{largesymbols}{"65}

\raggedright
\advance\textheight24pt

\begin{document}

\maketitle

This package provides some miscellaneous tools for mathematical 
accents, with the following goals in mind:
\begin{itemize}
\item Creating faked accents from non-accent symbols, like
$\accentset{\star}{s}$.
\item Grouping them, perhaps including actual accents, like 
$\hat{\accentset{\star}{\hat h}}$. That
has the side effect that multiple actual accents can be built
as well.
\item Putting them below the main symbol instead of above.
\end{itemize}

That is done so that the faked accents behave in quite similar fashion 
to actual accents do; i.e., the skew of the letter is taken into 
account (compare $\accentset{\ast}{d}$ with 
$\accentset{\ast}{h}$) and the sub and superscripts attached to 
it aren't misplaced (look carefully at $\accentset{\star}{f}_1^1$).

Release 1.1 included a few new features, some of them following 
suggestions by Donald Arseneau.  In particular, the |\underaccent| 
command has been fully reimplemented for the slant to be taken into 
account and the ``accent'' to be placed below the symbol can be 
anyone, not only real accents. This release just makes it compatible
with \textsf{amsmath} 2 with a quick fix.

\begin{decl}
|\DeclareMathAccent|
\end{decl}

This \LaTeXe{} command is reimplemented so that newly defined accents 
will incorporate the features of this package.  The standard accents 
are automatically redefined, including |\mathring| if you are using 
one of the latest \LaTeX{} releases.  However, both |\widetilde| and
|\widehat| remains untouched.  If you are using a non standard 
math encoding, the accents following the standard encoding names are 
rightly redefined, but new accents are not converted because 
\textsf{accents} is not aware of its existence.

See |fntguide.tex| in the \LaTeX{} standard distribution for a discussion on
 |\DeclareMathAccent|.

\begin{decl}
|\ring{<symbol>}|
\end{decl}

The accent in $\ring{x}$, which was available in this package (and
in fact in many others) before
the |\mathring| command was added to the \LaTeX{} kernel.

\begin{decl}
|\accentset{<accent>}{<symbol>}|
\end{decl}

Builds a faked accent, as for instance
\begin{sample}
|\accentset{\star}{d}|    \> $\accentset{\star}{d}$\\
|\accentset{\diamond}{h}| \> $\accentset{\diamond}{h}$\\
|\tilde{\accentset{\circ}{\phi}}| \>
    $\tilde{\accentset{\circ}{\phi}}$
\end{sample}
The |<accent>| is always in |\scriptscriptmode|; hence, using
|\accentset| in scripts won't give the desired  result. Of course,
if you use some faked accent many times, you can define:
\begin{verbatim}
\newcommand\starred[1]{\accentset{\star}{#1}}
\end{verbatim}
and |\starred| will become an accent, like |\hat|, |\tilde|, etc.

\begin{decl}
|\dddot  \ddddot|
\end{decl}

Two prefabricated faked accents: $\dddot{f}$ and $\ddddot{f}$.

\begin{decl}
|\underaccent{<accent>}{<symbol>}|
\end{decl}

This command puts the |<accent>| under the |<symbol>|. Both real
and faked accents are allowed. For instance,
\begin{sample}
|\underaccent{\hat}{x}|      \> $\underaccent{\hat}{x}$\\
|\underaccent{\bar}{\gamma}| \> $\underaccent{\bar}{\gamma}$\\
|\underaccent{\triangleright}{q}|  \>
   $\underaccent{\triangleright}{q}$\\
|\underaccent{\tilde}{\mathcal{A}}| \>
    $\underaccent{\tilde}{\mathcal{A}}$
\end{sample}

Many people likes using the wider version of the tilde accent as 
printed by the |\widetilde| command, i.e., $\widetilde{A}$ instead of 
$\tilde{A}$.  I find that aesthetically questionable, but anyway it
can be used under the symbol as well. Since |\widetilde| remains
untouched, you should define:
\begin{verbatim}
\DeclareMathAccent{\wtilde}{\mathord}{largesymbols}{"65}
\end{verbatim}
and write |\underaccent{\wtilde}{V}|, say (giving 
$\underaccent{\wtilde}{V}$). You may build an accent with
|\sim|, too. (The value for a wide hat is |"62|)

Sadly, \TeX{} lacks of a mechanism to place underaccents similar to 
that used in accents.  Letters have a large variety of shapes and 
finding an automatic adjusting is practically impossible.  Compare for 
instance the following letters:
\begingroup
\def\\{\underaccent{\bar}}%
$\\V$, $\\Q$, $\\p$, $\\q$, $\\f$,
$\\\beta$, $\\\gamma$, $\\{\mathcal{F}}$, $\\{\mathcal{A}}$
\endgroup
and you wil understand the difficulties.

\begin{decl}
|\undertilde{<symbols>}|
\end{decl}

This is the ``under'' version of |\widetilde| and like the latter is intended for constructions
involving several symbols. For instance:
\begin{sample}
|\undertilde{CV}| \> $\undertilde{CV}$
\end{sample}
Note that in this case no correction is made in the placement of the tilde.

\begin{decl}
|nonscript single|
\end{decl}

These package options are intended mainly for speeding up the 
typesetting of document.  The algorithm used here is recursive and 
very slow; although in fast processors that is not felt, in slow 
system that could be very annoying.

\begin{description}
\item[single] If you are interested only in |\accentset|.
 Accents are not reimplemented.
\item[nonscript] If you intend to use multiple accents in
 text and display modes only.
\end{description}

Macros are speeded up dramaticaly with both options; if your system is slow,
I commend using them in drafts and removing them for the final print.

Finally, some remarks:
\begin{itemize}
\item The package does not provides alternative accents for fonts
  lacking them. If you want an accented  |\mathcal| letter you had to
  write |\hat{\hat{\mathcal{A}}}| ($\hat{\hat{\mathcal{A}}}$).
\item |\mathbf{\hat{\hat h}}| gets the bold accent;
 |\hat{\hat{\mathbf{h}}}| not.
\item The symbol in |\accentset| must be a single symbol.
\item If you use \textsf{accents} with \textsf{amsmath} 2, you must
load \textsf{accents} after. Note that \textsf{amsmath} could redefine
some accents; in particular, if you experience problems with |\vec|
and you are using the standard math encodings, define:
\begin{verbatim}
\let\vec\relax
\DeclareMathAccent{\vec}{\mathord}{letters}{"7E}
\end{verbatim}
\end{itemize}

\end{document}

